---
title: "TaxSEA"
author: "Feargal Ryan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TaxSEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# TaxSEA: Taxon Set Enrichment Analysis

TaxSEA is an R package designed to enable rapid annotation of fold changes
or correlation coefficients in microbiome data. TaxSEA takes as input a
named vector of ranks (E.g. log2FC, Spearmans rho) and then uses a Kolmogorov-Smirnov test to identify if a particular taxa set is enriched
or depleted at one end of the distribution. For example, if there is a
depletion of butyrate producers, or enrichment of taxa previously found
to be associated with diabetes. TaxSEA does this by drawing upong multiple
published databases. 

## Taxon set database
TaxSEA utilizes taxon sets generated from five reference databases (**gutMGene**, **GMrepo v2**, **MiMeDB**, **mBodyMap**, **BugSigDB**). 

Please cite the appropriate database if using:

- Cheng et al. gutMGene: a comprehensive database for target genes of gut microbes and microbial metabolites Nucleic Acids Res. 2022.

- Dai et al. GMrepo v2: a curated human gut microbiome database with special focus on disease markers and cross-dataset comparison Nucleic Acids Res. 2022.

- Wishart et. al. MiMeDB: the Human Microbial Metabolome Database Nucleic Acids Res. 2023.

- Jin et al. mBodyMap: a curated database for microbes across human body and their associations with health and diseases. Nucleic Acids Res. 2022.

- Geistlinger et al. BugSigDB captures patterns of differential abundance across a broad range of host-associated microbial signatures. Nature Biotech. 2023. 

## Test data
The test data provided with TaxSEA consists of log2 fold changes comparing between healthy and IBD. The count data for this was downloaded from curatedMetagenomeData and fold changes generated with LinDA.
- Hall et al. A novel Ruminococcus gnavus clade enriched in inflammatory 
bowel disease patients** Genome Med. 2017 Nov 28;9(1):103.
- Pasolli et al. Accessible, curated metagenomic data through 
ExperimentHub. Nat Methods. 2017 Oct 31;14(11):1023-1024. doi: 10.1038/nmeth.4468.
- Zhou et al. LinDA: linear models for differential abundance analysis of microbiome compositional data. Genome Biol. 2022 Apr 14;23(1):95.


## Functions
- `get_taxon_sets(taxon)`: Retrieves taxon sets which contain a particular 
taxon for a list of taxon names.
- `get_ncbi_taxon_ids(taxon_names)`: Retrieves NCBI Taxonomy IDs for a 
list of taxon names.
- `TaxSEA(taxon_ranks, database = "All")`: Performs taxon set enrichment analysis.1
- `TaxSEA_barplot(taxsea_results, threshold = 0.2, custom_colors = NULL)`: Creates a bar plot of TaxSEA results.
- `TaxSEA_barcode(taxsea_results, taxon_ranks, taxon_sets, axis_limits = c(-7,7), n_to_plot = 10, boxplot = FALSE)`: Creates a barcode plot of 
TaxSEA results.

## Usage
```{r example}
library(TaxSEA)

# Retrieve taxon sets containing Bifidobacterium longum.
blong.sets <- get_taxon_sets(taxon="Bifidobacterium_longum")

# Run TaxSEA with test data provided
data("TaxSEA_test_data")
taxsea_results <- TaxSEA(taxon_ranks=TaxSEA_test_data)

#Enrichments among metabolite producers from gutMgene and MiMeDB
metabolites.df = taxsea_results$Metabolite_producers

#Enrichments among health and disease signatures from GMRepoV2 and mBodyMap
disease.df = taxsea_results$Health_associations

#Enrichments amongh published associations from BugSigDB
bsdb.df = taxsea_results$BugSigdB

```

#### Input 
All that is required for TaxSEA is a named vector of log2 fold changes between groups. This should be for all taxa tested, and not limited to only a pre-defined set (e.g. do not use a threshold for significance or remove any taxa). See example below for format. TaxSEA will lookup and convert taxon names to NCBI taxonomic identifiers. TaxSEA stores a commonly used identifiers internally and so will only look up whatever is not covered to save time. 
```{r output}
head(sample(TaxSEA_test_data),4)

```

#### Output
The output is a list of three dataframes providing enrichment results for metabolite produers, health/disease associations, and published signatures from BugSigDB.
Each dataframe has 5 columns
- taxonSetName - The name of the taxon set tested
- NES - Normalized enrichment score. This is simply the median fold change across the entire set
- P value - Kolmogorov-Smirnov test P value.
- FDR - P value adjusted for multiple testing. 
- TaxonSet - Returns list of taxa in the set to show what is driving the signal


#### BugSigDB
The format of BugSigDB is that each publication is entered as a "Study", and within this there is different experiments and signatures. 
Should users wish to find out more information about the signatures. For 
example different signatures may be the taxa that are increased or 
decreased. 
from the BugSigDB, they can do so by querying that database.  

```{r bugsigdb, eval=FALSE}
library(bugsigdbr) #This package is installable via Bioconductor
bsdb <- importBugSigDB() #Import database 

#E.g. if the BugSigDB identifier you found enriched was #bsdb:74/1/2_obesity:obese_vs_non-obese_DOWN
#This is Study 74, Experiment 1, Signature 2
bsdb[bsdb$Study=="Study 74" & 
     bsdb$Experiment=="Experiment 1" & 
     bsdb$Signature=="Signature 2",]



```




#### TaxSEA database with other enrichment tools
The TaxSEA function by default uses the Kolmogorov Smirnov test. This was the original test used for gene set enrichment analysis, however subsequent methods have been developed based on alternative approaches. One particularly popular package is fast gene set enrichment analysis (fgsea) which is based upon a permutation type approach. The TaxSEA database is compatible with fgsea as long as the input identifiers are NCBI IDs. If they are not currently in that format we provide a function for converting. See example code below. 

```{r fgsea,eval=FALSE}
library(fgsea) #This package is installable via Bioconductor

#Convert input names to NCBI taxon ids
names(TaxSEA_test_data) = get_ncbi_taxon_ids(names(TaxSEA_test_data))
TaxSEA_test_data = TaxSEA_test_data[!is.na(names(TaxSEA_test_data))]

#Run fgsea
fgsea_results <- fgsea(TaxSEA_db, TaxSEA_test_data, minSize=5, maxSize=500)
```



```{r sessioninfo}
sessionInfo()
```

