---
title: "Single-sample enrichment with ssTaxSEA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{single_sample_enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(TaxSEA)
```

## Why per-person enrichment?

Standard TaxSEA works at the group level. You run a differential abundance
analysis, get a fold change for each taxon, and ask whether biologically
related taxa shift together. That is powerful for characterising what differs
between groups.

But sometimes the more interesting question is about individuals:

- Which patients in my IBD cohort show an oral bacteria signature?
- Does this person's microbiome look more "disease-like" for a particular trait?
- Can I stratify my cohort by enrichment patterns rather than by diagnosis alone?

Group-level enrichment cannot answer these questions because it collapses all
individuals into a single summary statistic. `ssTaxSEA()` produces a score for
every sample, for every taxon set.

## How it works

For each sample, the counts are centered log-ratio (CLR) transformed, and the
score for a taxon set is simply the **mean of the CLR values of the set's
members**:

$$score_{S,j} = \frac{1}{|S|} \sum_{i \in S} clr_{ij}$$

That is the whole method. There is no ranking, no running-sum statistic and no
cohort-relative standardisation.

Two details are worth stating explicitly.

**The CLR is computed across all taxa you supply, before any subsetting to set
members.** The CLR divides each taxon by the geometric mean of the sample, so
that denominator has to represent the whole community. If you were to subset the
matrix to set members first and transform afterwards, the denominator would
depend on which sets you happened to be testing — and sets with no real signal
would pick up an apparent one. `ssTaxSEA()` does this in the right order; the
point matters if you are ever tempted to hand-roll the calculation.

**The score is a within-sample quantity.** It does not reference the rest of the
cohort at all, which means `ssTaxSEA()` is well defined for a single sample.

## Usage

### From a count matrix

Rows are taxa, columns are samples, and row names are taxon names in the same
format standard TaxSEA expects.

```{r matrix-example}
set.seed(42)
counts <- matrix(rpois(30 * 8, lambda = 10), nrow = 30, ncol = 8)
rownames(counts) <- paste0("Taxon_", seq_len(30))
colnames(counts) <- paste0("Sample_", seq_len(8))

# Plant a signal: raise one set's members in the last four samples
counts[1:6, 5:8] <- counts[1:6, 5:8] * 5

my_sets <- list(
  elevated_set = paste0("Taxon_", 1:6),
  control_set  = paste0("Taxon_", 20:27)
)

scores <- ssTaxSEA(counts, custom_db = my_sets, min_set_size = 3)
dim(scores)
round(scores, 2)
```

The returned object is a plain numeric matrix with **taxon sets as rows and
samples as columns**. The planted signal is visible as higher scores in the last
four samples of `elevated_set`:

```{r matrix-signal}
rowMeans(scores[, 5:8]) - rowMeans(scores[, 1:4])
```

### A single sample

Because the score never references the cohort, one sample is enough:

```{r single-sample}
ssTaxSEA(counts[, 1, drop = FALSE], custom_db = my_sets, min_set_size = 3)
```

### With the built-in database

Leave `custom_db` unset to use the built-in TaxSEA sets. Row names are mapped to
NCBI taxonomy IDs, exactly as in standard `TaxSEA()`.

```{r builtin, eval = FALSE}
scores <- ssTaxSEA(counts)
dim(scores)
```

### From a TreeSummarizedExperiment

`ssTaxSEA()` accepts `SummarizedExperiment` and `TreeSummarizedExperiment`
objects directly, extracting the first assay and using `rownames()` as taxon
identifiers.

```{r tse, message = FALSE, eval = requireNamespace("SummarizedExperiment", quietly = TRUE)}
library(SummarizedExperiment)

se <- SummarizedExperiment(assays = list(counts = counts))
scores_se <- ssTaxSEA(se, custom_db = my_sets, min_set_size = 3)

# Identical to the matrix result
all.equal(scores, scores_se)
```

## Interpreting the scores

**Compare across samples, within a set.** A higher score for `elevated_set` in
sample 6 than in sample 2 means the set's members make up more of sample 6's
community. This is the comparison the score is built for.

**Do not compare across sets.** A set of abundant taxa scores higher than a set
of rare taxa in every sample, regardless of biology, because its members are
simply more abundant. In the example above the two sets sit at different
baselines for exactly this reason:

```{r baselines}
rowMeans(scores)
```

For heatmaps and clustering, center or scale the rows first so that you are
looking at variation between samples rather than differences in baseline
abundance:

```{r centering}
scaled <- t(scale(t(scores)))
round(scaled, 2)
```

**No p-values are returned.** The score is a descriptive statistic, and a
p-value computed within a single sample would mostly be telling you whether the
set's members are more abundant than the average taxon — a property of the set,
not of the sample. To test a hypothesis, take one row of the matrix and compare
it across samples with a test suited to your design:

```{r testing}
group <- rep(c("control", "case"), each = 4)
wilcox.test(scores["elevated_set", ] ~ group)$p.value
```

**Remember that the data are compositional.** Because CLR values within a sample
are constrained, raising one set's members slightly lowers everything else. Small
shifts in sets you believe to be null are expected, and with enough samples they
can reach statistical significance while being negligible in size. Report effect
sizes alongside p-values rather than relying on p-values alone.

## Practical considerations

**The pseudocount matters.** Zeros cannot be log-transformed, so `pseudocount`
(default 0.5) is added to every value before the transform. Adding it to all
values rather than only to zeros keeps the transform monotonic in the counts.
The choice does affect the scores, so report the value you used.

```{r pseudocount}
p1 <- ssTaxSEA(counts, custom_db = my_sets, min_set_size = 3, pseudocount = 1)
round(p1["elevated_set", ] - scores["elevated_set", ], 3)
```

**Supply counts, not proportions.** Adding a pseudocount of 0.5 to values that
sum to 1 would overwhelm the data completely, so `ssTaxSEA()` refuses input whose
columns sum to 1. If you only have relative abundances, rescale them to a
count-like scale first.

**BugSigDB is excluded.** `ssTaxSEA()` uses the built-in TaxSEA database
(GMRepoV2, MiMeDB, gutMGene, mBodyMap, BacDive) but does not load BugSigDB. This
keeps the function fast and avoids the `bugsigdbr` dependency.

## When to use ssTaxSEA vs TaxSEA

| | **TaxSEA** | **ssTaxSEA** |
|---|---|---|
| **Input** | Named vector of fold changes or correlations | Count matrix (taxa x samples) or TreeSummarizedExperiment |
| **Output** | One enrichment result per taxon set | One score per sample per taxon set |
| **Question** | What trait-based shifts characterise the difference between groups? | Which individual samples show a given enrichment pattern? |
| **Use case** | Interpreting DA results, contextualising group differences | Patient stratification, individual-level phenotyping |
| **Significance** | p-values and FDR per set | none; test scores across samples yourself |

In practice you might use both: run `TaxSEA()` to identify which taxon sets are
enriched at the group level, then `ssTaxSEA()` to see how those enrichments
distribute across individuals.

## Downstream analysis

The score matrix is designed to slot into standard workflows:

- Heatmaps of per-sample enrichment patterns (scale the rows first)
- Correlation with clinical metadata
- Group comparisons on a single set's scores
- Clustering or ordination to identify patient subgroups
- Regression or survival analysis using scores as predictors

```{r sessioninfo}
sessionInfo()
```
