---
title: "TaxSEA"
author: "Feargal Ryan"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{TaxSEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# TaxSEA: Taxon Set Enrichment Analysis

TaxSEA is an R package designed to enable rapid interpretation of 
differential abundance analysis  or correlation analysis output for 
microbiota data. 
TaxSEA takes as input a vector of genus or species names and a rank. 
For example log2 fold changes, or Spearman's rho. TaxSEA then uses
a Kolmogorov-Smirnov test to identify if a particular group of species
or genera (i.e. a set of taxa such as butyrate producers) are skewed to 
one end of the distribution . 

Simply put, TaxSEA allows users to rapidly go from a list of species to 
which metabolite producers are altered, and if the findings are similar to
a previously published study. 


## Installation

To install the latest version of TaxSEA from Bioconductor:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("TaxSEA")
```


## Taxon set database
TaxSEA utilizes taxon sets generated from five reference databases 
(**gutMGene**, **GMrepo v2**, **MiMeDB**, **mBodyMap**, **BugSigDB**). 

Please cite the appropriate database if using:

- Cheng et al. gutMGene: a comprehensive database for target genes of gut 
microbes and microbial metabolites Nucleic Acids Res. 2022.

- Dai et al. GMrepo v2: a curated human gut microbiome database with 
special focus on disease markers and cross-dataset comparison 
Nucleic Acids Res. 2022.

- Wishart et. al. MiMeDB: the Human Microbial Metabolome Database 
Nucleic Acids Res. 2023.

- Jin et al. mBodyMap: a curated database for microbes across human body 
and their associations with health and diseases. Nucleic Acids Res. 2022.

- Geistlinger et al. BugSigDB captures patterns of differential abundance 
across a broad range of host-associated microbial signatures. 
Nature Biotech. 2023. 

## Test data
The test data provided with TaxSEA consists of log2 fold changes comparing 
between healthy and IBD. The count data for this was downloaded from 
curatedMetagenomeData and fold changes generated with LinDA.

- Hall et al. A novel Ruminococcus gnavus clade enriched in inflammatory 
bowel disease patients Genome Med. 2017 Nov 28;9(1):103.
- Pasolli et al. Accessible, curated metagenomic data through 
ExperimentHub. Nat Methods. 2017 Oct 31;14(11):1023-1024. 
doi: 10.1038/nmeth.4468.
- Zhou et al. LinDA: linear models for differential abundance analysis of 
microbiome compositional data. Genome Biol. 2022 Apr 14;23(1):95.


## Functions
- `get_taxon_sets(taxon)`: Retrieves taxon sets which contain a particular 
taxon for a list of taxon names.
- `get_ncbi_taxon_ids(taxon_names)`: Retrieves NCBI Taxonomy IDs for a 
list of taxon names.
- `TaxSEA(taxon_ranks, database = "All")`: Taxon set enrichment analysis.


## Usage
### Retrieve sets containing a particular taxon
```{r example1}
library(TaxSEA)

# Retrieve taxon sets containing Bifidobacterium longum.
blong.sets <- get_taxon_sets(taxon="Bifidobacterium_longum")
```

### Running an enrichment analysis
All that is required for TaxSEA is a named vector of log2 fold changes 
between groups for species or genera. TaxSEA will not work for ranks higher
than species or genus. The input should be for all taxa tested, and not 
limited to only a pre-defined set (e.g. do not use a threshold for 
significance or remove any taxa). See example below for format. TaxSEA will
lookup and convert taxon names to NCBI taxonomic identifiers. TaxSEA stores
commonly observed identifiers internally to save time. 

TaxSEA can also utilise custom databases which should be a named list
of taxon sets. In this case the ID conversion is disabled and it is
expected that the input names and database names will be in the same
format

Input IDs should be in the format of like one of the following
- Species name. E.g. "Bifidobacterium longum", "Bifidobacterium_longum"
- Genus name. E.g. "Bifidobacterium"
- NCBI ID E.g. 216816


```{r input_data}
#Input IDs with the full taxonomic lineage should be split up. E.g.
x <- paste0(
  "d__Bacteria.p__Actinobacteriota.c__Actinomycetes.",
  "o__Bifidobacteriales.f__Bifidobacteriaceae.g__Bifidobacterium")
x = strsplit(x,split="\\.")[[1]][6]
x = gsub("g__","",x)
print(x)

## Example test data
library(TaxSEA)
data(TaxSEA_test_data)
head(sample(TaxSEA_test_data),4)

```



### Run TaxSEA with test data
```{r example2}
data("TaxSEA_test_data")
taxsea_results <- TaxSEA(taxon_ranks=TaxSEA_test_data)

#Enrichments among metabolite producers from gutMgene and MiMeDB
metabolites.df = taxsea_results$Metabolite_producers

#Enrichments among health and disease signatures from GMRepoV2 and mBodyMap
disease.df = taxsea_results$Health_associations

#Enrichments amongh published associations from BugSigDB
bsdb.df = taxsea_results$BugSigdB

```


#### Output
The output is a list of three dataframes providing enrichment results for 
metabolite produers, health/disease associations, and published signatures 
from BugSigDB.
Each dataframe has 5 columns
- taxonSetName - The name of the taxon set tested
- median_rank - The median rank of set members
- P value - Kolmogorov-Smirnov test P value.
- FDR - P value adjusted for multiple testing. 
- TaxonSet - Returns list of taxa in the set to show what is driving the 
signal


#### BugSigDB
The format of BugSigDB is that each publication is entered as a "Study",
and within this there is different experiments and signatures. 
Should users wish to find out more information about the signatures,
they can do so by querying that database. 


```{r bugsigdb}
library(bugsigdbr) #This package is installable via Bioconductor
bsdb <- importBugSigDB() #Import database 

#E.g. if the BugSigDB identifier you found enriched was 
#bsdb:74/1/2_obesity:obese_vs_non-obese_DOWN
#This is Study 74, Experiment 1, Signature 2
bsdb[bsdb$Study=="Study 74" & 
     bsdb$Experiment=="Experiment 1" & 
     bsdb$Signature=="Signature 2",]

```

#### TaxSEA database with other enrichment tools
The TaxSEA function by default uses the Kolmogorov Smirnov test and the 
original idea was inspired by gene set enrichment analyses from RNASeq.
Should users wish to use an alternative gene set enrichment analysis tool
the database is formatted in such a way that should be possible. See below
for an example with fast gene set enrichment analysis (fgsea). 

```{r fgsea}
library(fgsea) #This package is installable via Bioconductor
data("TaxSEA_test_data")
data("TaxSEA_db")
#Convert input names to NCBI taxon ids
names(TaxSEA_test_data) = get_ncbi_taxon_ids(names(TaxSEA_test_data))
TaxSEA_test_data = TaxSEA_test_data[!is.na(names(TaxSEA_test_data))]

#Run fgsea
fgsea_results <- fgsea(TaxSEA_db, TaxSEA_test_data, minSize=5, maxSize=500)
```

```{r sessioninfo}
sessionInfo()
```


