---
title: "BacDive physiology sets"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

TaxSEA includes taxon sets built from [BacDive](https://bacdive.dsmz.de), the
Bacterial Diversity Metadatabase. Each set groups species that share a
measured physiological property: they tolerate oxygen, stain Gram-negative,
ferment glucose, reduce nitrate, grow above 45 °C, and so on. Testing these
sets asks whether a shift in your data is a shift in *what the bacteria can
do*, not just *which bacteria are there*.

Since TaxSEA 1.5.5 these sets are rebuilt from a complete harvest of BacDive
and contain measured laboratory data only.

## What is in them

```{r}
library(TaxSEA)
data(TaxSEA_db)

bacdive <- TaxSEA_db[startsWith(names(TaxSEA_db), "BacDive_")]
length(bacdive)                                   # number of sets
length(unique(unlist(bacdive, use.names = FALSE))) # taxa covered
```

Set names read `BacDive_<Family>_<value>`, optionally followed by an assay
context and `_negative`:

```{r}
family <- sub("^BacDive_([^_]+)_.*", "\\1", names(bacdive))
sort(table(family), decreasing = TRUE)
```

| Family | What it describes | Examples |
|---|---|---|
| `Oxygen` | oxygen tolerance | `BacDive_Oxygen_anaerobe`, `BacDive_Oxygen_facultative_anaerobe` |
| `Gram`, `Motility`, `Spore`, `Shape` | cell properties | `BacDive_Gram_negative`, `BacDive_Spore_yes`, `BacDive_Shape_coccoid` |
| `Temp`, `NaCl`, `pH`, `GC` | growth range and genome GC, from numeric measurements | `BacDive_Temp_thermophile`, `BacDive_NaCl_halotolerant` |
| `Uses` | substrate use, split by assay | `BacDive_Uses_glucose_fermentation`, `BacDive_Uses_nitrate_reduction` |
| `Enzyme` | enzyme activities, including hydrolysis | `BacDive_Enzyme_catalase`, `BacDive_Enzyme_esculin_hydrolysis` |
| `Prod`, `Produces`, `Test` | metabolites produced; classic biochemical tests | `BacDive_Prod_indole`, `BacDive_Produces_vitamin_B12` |
| `Observation`, `Murein`, `Nutrition`, `Tolerance`, `PathogenHuman` | quinones, peptidoglycan type, nutrition type and more | `BacDive_Observation_menaquinone` |

### Assay context matters

Growing on a substrate, fermenting it, producing acid or gas from it,
reducing it and respiring it are different physiology, so they are
different sets:

```{r}
grep("^BacDive_Uses_nitrate", names(bacdive), value = TRUE)
```

### Negative sets

A species that was tested and found negative is not the same as one that was
never tested. Sets ending in `_negative` hold the tested negatives, so you can
ask, for example, whether catalase-negative species shift:

```{r}
lengths(bacdive[c("BacDive_Enzyme_catalase", "BacDive_Enzyme_catalase_negative")])
```

## Using them

BacDive results come back in their own table:

```{r}
data(TaxSEA_test_data)
res <- TaxSEA(taxon_ranks = TaxSEA_test_data)
head(res$BacDive_bacterial_physiology[, c("taxonSetName",
                                         "median_rank_of_set_members",
                                         "FDR")])
```

To see which BacDive sets a species belongs to:

```{r}
ecoli <- get_taxon_sets("Escherichia_coli")
head(grep("^BacDive_Oxygen|^BacDive_Gram|^BacDive_Uses_nitrate", ecoli,
          value = TRUE))
```

Species names can be given with spaces or underscores, and former names of
reclassified species work too (*Bacteroides vulgatus* and *Phocaeicola
vulgatus* both resolve). NCBI taxonomy IDs can be given directly.

## How many strains stand behind a set

Every set ships with its provenance: which BacDive field it came from and how
many species and strains support it.

```{r}
prov <- read.delim(system.file("extdata", "BacDive_set_provenance.tsv",
                               package = "TaxSEA"))
head(prov[order(-prov$n_taxa),
          c("set_name", "family", "n_species", "n_strains_total")])
```

`BacDive_build_info.tsv` in the same folder records when BacDive was
harvested and which harvester release built the sets.

## Things to keep in mind

- **BacDive is silent for some well-known gut taxa.** It holds no oxygen
  tolerance record for *Faecalibacterium prausnitzii* or *Roseburia
  intestinalis*, so they are not in the oxygen sets. Earlier TaxSEA releases
  listed them from a curated table; the sets now reflect BacDive only.
- **Sets are correlated.** A species that ferments glucose often ferments
  mannose, so several substrate sets can be significant for the same
  underlying reason.
- **Very small sets are left out.** Only sets with at least 3 members are
  included.

## Changes in TaxSEA 1.5.5

Set names changed. The most common old names map as follows:

| Before 1.5.5 | From 1.5.5 |
|---|---|
| `BacDive_facultative anaerobe` | `BacDive_Oxygen_facultative_anaerobe` |
| `BacDive_anaerobe` | `BacDive_Oxygen_anaerobe` |
| `BacDive_rod-shaped` | `BacDive_Shape_rod` |
| `BacDive_Utilizes_nitrate` | `BacDive_Uses_nitrate_reduction` (nearest equivalent) |
| `BacDive_Enzyme_alkaline phosphatase` | `BacDive_Enzyme_alkaline_phosphatase` |

The sets are built by
[bacdive_harvester](https://github.com/feargalr/bacdive_harvester), which
documents every curation decision.

Please cite BacDive if you use these sets: Schober et al. BacDive in 2025: the
core database for prokaryotic strain data. *Nucleic Acids Res.* 2025.
